---
title: 'BIOS 735 Final Project: Car Accident Severity'
author: "Marissa Ashner, Yen Chang, Marco Chen, Weifang Liu, Yi Tang, Joyce Yan"
date: "4/27/2020"
output:
  html_document:
    df_print: paged
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 

# Project Aim 

The aim of this project is to use Car Accident Data from North Carolina in 2016-2018 to determine which factors have the most influence on the severity of car accidents in the state. Then, we aim to use these factors to build a model to predict the severity of future car accidents in North Carolina and compare the accuracy with 2019 data.

# R-Package

Our R package `name` contains ...

```{r}
## load package here 
```

# Data

```{r}
data <- read.csv("NC_Accidents.csv")
```

Variables: 
1. ID -- no duplicated IDs, all unique 
2. Source (API that reported accident) -- remove
3. TMC -- remove
4. Severity -- response 
5. Start Time -- keep the hour and minute of the day?
6. End Time 
7. Start Lat -- remove 
8. Start Lng -- remove
9. End Lat -- remove
10. End Lng -- remove
11. Distance (Length of road extent affected by accident)
12. Description (in Natural Language) -- remove for now?
13. Number (Street) -- remove
14. Street (Name) -- remove
15. Side (of Street -- R/L)
16. City -- remove for now? or aggregate rural/urban?
17. County -- remove for now? or aggregate rural/urban?
18. State -- remove
19. Zipcode -- remove 
20. Country -- remove
21. Timezone -- remove 
22. Airport Code -- remove
23. Weather Timestamp -- remove
24. Temperature 
25. Wind Chill 
26. Humidity (%)
27. Pressure (Air, in inches)
28. Visibility (in miles)
29. Wind Direction 
30. Wind Speed (mph) 
31. Precipitation (in inches)
32. Weather Condition -- aggregate?
33. Amenity -- remove
34. Bump (presence of speed bump in nearby location)
35. Crossing (presence of crossing nearby)
36. Give_Way (presence of give_way sign nearby)
37. Junction (presence of junction nearby)
38. No_Exit (presence of no_exit sign nearby)
39. Railway (presence of railway nearby)
40. Roundabout 
41. Station 
42. Stop (sign)
43. Traffic_Calming 
44. Traffic_Signal 
45. Turning_Loop
46. Sunrise_Sunset -- pick one of these last four 
47. Civil_Twilight
48. Nautical_Twilight
49. Astronomical_Twilight


# Methods 

## Proportional Odds Model 

## Random Forest

# Results 

## Proportional Odds Model 

## Random Forest

## Method Comparison 

# Discussion 

# Data References 

https://www.kaggle.com/sobhanmoosavi/us-accidents

Moosavi, Sobhan, Mohammad Hossein Samavatian, Srinivasan Parthasarathy, and Rajiv
  Ramnath. “A Countrywide Traffic Accident Dataset.”, 2019.
  
Moosavi, Sobhan, Mohammad Hossein Samavatian, Srinivasan Parthasarathy, Radu
  Teodorescu, and Rajiv Ramnath. "Accident Risk Prediction based on Heterogeneous
  Sparse Data: New Dataset and Insights." In proceedings of the 27th ACM SIGSPATIAL
  International Conference on Advances in Geographic Information Systems, ACM, 2019.

